---
title: "Lecture 19: A Few R Demos, Interactions and Collinearity"
subtitle: "BIOS 600 - Spring 2026"
institute: "UNC - Chapel Hill"
logo: "images/logo_600.png"
format:
  revealjs:
    css: reveal-custom.css
execute:
  freeze: auto
---

## Announcements

- Exam Corrections due **Thursday 11/20 at 11:59pm**.

- Lab 8 due this **Friday at 11:59pm**. 

## Overview

- Some R demos, helpful project things

- Interpreting categorical variables, plus releveling factors

- Interactions & how to interpret interaction terms

- Checking for multicollinearity


## Computational setup

```{r}
#| echo: true
#| message: false
#| warning: false
library(tidyverse)
library(tidymodels)
library(kableExtra)
library(car) # for VIF
```


## R Demo: merging datasets

- Merging datasets could be helpful for your project!

- Let's say you have two data sets..

```{r}
# Create two sample data frames
df1 <- data.frame(
  ID = c(1, 2, 3, 4),
  Name = c("Alice", "Bob", "Charlie", "David")
)

df2 <- data.frame(
  ID = c(1, 3, 5),
  City = c("New York", "London", "Paris")
)
```

```{r}
#| echo: true
df1

df2
```

## R Demo: Merging datasets

- As long as you have an identifying column (e.g. studyid, patient number, observation number etc.), you can merge using `left_join()`:

```{r}
#| echo: true
# The 'by' argument specifies the 
# common column(s) to join on.
merged_df <- df1 |>
  left_join(df2, by = "ID")

merged_df 
```
## R Demo: Dropping NA values

- We haven't talked about missing data much, but it may be simplest to conduct a "complete case analysis" where you only perform an analysis on complete cases:

```{r}
#| echo: true
merged_df |>
  drop_na()
```

## R Demo: `write.csv()`

- You can save your cleaned/merged dataset at any point using `write.csv()`

```{r}
#| echo: true
#| eval: false

write.csv(merged_df, file = "data/data_clean.csv", 
          row.names = F) 
# do not create a column of row names
```

- Note that you may need to include `row.names = F` so that R does not create an additional column of row numbers.


## Categorical variables from last time...

```{r}
#| echo: false
#| warning: false
#| message: false
cdc <- read.csv("https://karamccor.github.io/b6002/labs/data/cdc_cleaned.csv")
```

Consider the model

::: poll
$$Obesity_i = \beta_0 + \beta_1(HDI == Middle)_i + $$

$$ \beta_2(HDI == BottomTen)_i + \epsilon_i$$
:::

The parameter interpretations are below.

-   $\beta_0$ represents the expected obesity percentage for a state with 0 for the two dummy variables. That is, in the top ten HDI

-   $\beta_1$ represents the expected difference in obesity percentage for a state in the middle HDI category, **compared to the top ten**

-   $\beta_2$ represents the expected difference in obesity percentage for a state in the bottom ten HDI category, **compared to the top ten**

## Fitting the model

```{r echo = T}
fit2 <- lm(Obesity ~ HDI, data = cdc)
tidy(fit2) |>
  kable(digits = 3)
```


- Wait a minute! R automatically assigned the lowest level of HDI as the reference level. This is because R assigns factor levels to characters **alphabetically**, so `HDIBottom Ten` was before `HDIMiddle` and `HDITop Ten`.

- Let's see how we can change the reference level. But before doing that...

## Writing out the model

::: callout-tip

## Writing out the model

- Write out the new model that has been fit. (i.e. $Obesity_i = \ldots$). Do we include an error term? Why or why not?

:::

## Changing the reference level



```{r}
#| echo: true
cdc_new <- cdc |>
  mutate(HDI_relevel = fct_relevel(HDI, "Top ten"))
```

- This reorders the factor levels so that "`Top ten`" becomes the first (reference) level. 

- Now, when we run the regression, the model will automatically use `Top ten` as the baseline. 

## Re-running the model

```{r}
#| echo: true
fit2_relevel <- lm(Obesity ~ HDI_relevel, 
                   data = cdc_new)
tidy(fit2_relevel) |>
  kable(digits = 3)
```

::: callout-tip
## Interpretation and Model

- What is the interpretation of $\hat\beta_2 = 3.283$ from the output above?


:::

## Interactions {.smaller}

-   Sometimes, the relationship between one predictor and the outcome depends on the value of another predictor variable. For example, the effect of exercise % on obesity may be different for smokers vs. non-smokers. To model such a relationship, we create an **interaction term**.

-   This is created simply by multiplying two predictors $x_1$ and $x_2$ to create a new predictor, $x_1x_2$. When interaction terms are in a model, interpretations can become tricky.

## Interactions

Let's consider a model with **main effects** of exercise and smoking and an **interaction term** between them:

$$Obesity_i = \beta_0 + \beta_1 Exercise + \beta_2 Smoking$$

$$+ \beta_3 (Exercise_i \cdot Smoking_i) + \epsilon_i$$

What is the expected change in obesity % given a one percentage point increase in exercise %?

## Interactions model

```{r}
#| echo: true
fit_int <- lm(Obesity ~ Exercise + 
                Smoking + 
                Exercise*Smoking, 
                   data = cdc)
tidy(fit_int) |>
  kable(digits = 3)
```

**Interpretation**: If exercise increases by 1 percentage point, the expected change in obesity % depends on smoking %, and is: -.358 + .003(Smoking%)

## More interpretation

So, 

- When Smoking = 0%, the slope of Exercise is -0.358 (exercise is assoc. w/ lower obesity).

- When Smoking = 50%, slope = -0.358 + 0.003(50) = -0.208 (still negative association, but smaller in magnitude).

**However**: The interaction term is not significant. So we should remove the interaction term and just use the model with main effects only. 

## The math behind it

- How did we get that interpretation? Here's the math behind it...

For a given exercise (ex) and smoking (sm) percentage, our predicted obesity (ob) percentage is

$$\hat {ob}_i = \hat \beta_0 + \hat \beta_1 ex_i + \hat \beta_2 sm_i + \hat \beta_3 ex_i sm_i$$

and for a state at the same smoking % but 1 percentage point higher in exercise %, the predicted obesity percentage is

$$\hat {ob}_{i'} = \hat \beta_0 + \hat \beta_1 (ex_i + 1) + \hat \beta_2 sm_i + \hat \beta_3 (ex_i + 1) sm_i$$

$$ = \hat \beta_0 +  \hat \beta_1 ex_i + \hat \beta_1 + \hat \beta_2 sm_i + \hat \beta_3 ex_i sm_i + \hat \beta_3 sm_i$$

## The math behind it

- Subtracting, we have $\hat ob_{i'} - \hat ob_i= \hat \beta_1 + \hat \beta_3 sm_i$, which is the expected change in obesity % for a 1% change in exercise %.

- **Takeaway**: In the interactions model, the relationship of exercise and obesity depends on the level of smoking in that state.

::: callout-tip
## Quick review

Take a few minutes to review the math in small groups.

:::

## Interactions {.smaller}

-   Luckily, interpretation of interaction terms with categorical predictors is easier than with continuous predictors.

-   Since categorical predictors are based on dummy variables, they can only take on the values of 0 or 1 in the model.

-   Again, an interaction effect implies that the regression coefficient for an explanatory variable would change depending on the value of another predictor (for instance, the relationship between exercise and obesity might depend on whether a state is in High, Medium, or Low HDI groups).


## Multicollinearity

-   One common problem in multiple regression is **multicollinearity**, which occurs when multiple highly correlated variables are used as predictors.

- **Collinearity** simply refers to two variables behind highly correlated.

-   In this case, the model can **become unstable** (often seen as standard errors that get huge and lead to huge confidence interval estimates), and it can be difficult to assess the relationships of the predictors.

## Diagnosing multicollinearity {.smaller}

If nothing is significant when you "expect" something to be, we have some clues:

-   Individual predictors are significant in simple linear regression models,

-   but standard errors and interval estimates are huge,

-   and the overall F test is significant

A **significant overall F test with no significant individual variable test** is a typical sign of collinearity. We can check out the correlations among the three predictors.

## Checking VIF

- The common way to check for multicollinearity is through VIF values. VIF values indicate the extent of multicollinearity.

- A VIF value of 1 suggests no correlation with other predictors.

- Values between 1 and 5 generally indicate moderate correlation.

- Values above 5 (or sometimes 10, depending on the context) are often considered problematic, indicating severe multicollinearity and potential issues with coefficient stability and interpretation. 

## VIF


- Let's say we had the following model:

```{r}
#| echo: true
fit4 <- lm(Obesity ~ Exercise + Smoking + Over65, 
           data = cdc)
vif_values <- vif(fit4) # car package
vif_values
```
- In this model, none of the VIFs are above 5 (in fact, they're much closer to 1), so multicollinearity is not an issue here. 

## Recap

- Some R demos: merging, dropping NA, `write.csv()`, re-leveling a factor

- Interactions, interpretations, and the math behind it all

- Collinearity

## Next class

- Logistic regression!
