---
title: "Lecture 17: Introduction to linear regression"
subtitle: "BIOS 600 - Spring 2026"
institute: "UNC - Chapel Hill"
logo: "images/logo_600.png"
format:
  revealjs:
    css: reveal-custom.css
execute:
  freeze: auto
---

## Announcements

- No HW until next week.

- Lab 7 (new lab) today - due Friday April 10th

## Uploading an excel file

- You can upload `.xlsx` files using the **readxl** package. 

```{r}
#| echo: true
#| eval: false
# Install (only once)
install.packages("readxl")

# Load each time you start R
library(readxl)

# read in data

# Example: import a dataset called "biostats_data.xlsx"
data <- read_excel("data/biostats_data.xlsx")

```

- Let me or the TAs know if you need help reading in a data file. 

## Topics covered in lab today

- In lab today you'll hear about two additional concepts
  - Partial correlation
  - Spearman's correlation coefficients

## Partial correlation

- Let's say I have an example of 100 adults, I have data on:
  - Age (in years)
  - Cognitive function score - ranging from 0-100
  - Systolic blood pressure (SBP)

- **Is higher SBP is associated with lower cognitive scores?**

- We anticipate
  - Older individuals have lower cognitive function scores
  - Older individuals have higher SBP 
  
## Partial correlation

- Age could be a **confounder** 
  - Age is associated with the exposure (SBP)
  - Age is associated with the outcome (cognitive function)

- If I look at the association between SBP and cognitive function I might get a biased estimate
  - Could look like two things are related when they aren't
  - Could look larger or smaller than it really is

## Partial correlation

- We'll get into more common methods (and a bit more detail on these ideas on Thursday)

- But **partial correlation** is one approach

- Isolates the direct association, preventing misleading results caused by confounding variables

- Partial correlation removes age's influence from both variables, then asks - **what's left**

```{r, echo=FALSE, warning=FALSE, message=FALSE}
library(tidyverse)
library(knitr)
library(ppcor)
# Simulate data
set.seed(7)
n <- 100

age <- round(rnorm(n, mean = 50, sd = 12))
sbp <- round(105 + 0.65 * (age - 50) + rnorm(n, 0, 9))
cog <- round(82 - 0.40 * (age - 50) - 0.05 * (sbp - 130) + rnorm(n, 0, 7))

study <- data.frame(id = 1:n, age, sbp, cog)
```

## Partial correlation

Here's the example dataset, with age, sbp and cognitive function

```{r}
head(study)
```

## Partial correlation

Pearson correlation

```{r}
cor(study$sbp, study$cog)
```

Partial correlation 

```{r}
pcor.test(study$cog,study$sbp,study$age)
```

## Spearman example 

- Pearson is parameteric - using magnitude and values - hen x goes up, does y tend to go up?" — it doesn't care how much. Pearson cares about the exact magnitudes

- Spearman is using ranks - when x goes up, does y tend to go up — it doesn't care how much
Monotonic vs. Linear: A linear relationship is monotonic (constantly increasing/decreasing), but a monotonic relationship is not necessarily linear (e.g., exponential growth). Spearman detects both.
Rank Transformation: By converting raw data to ranks, it assesses if ranks align, ignoring the magnitude of differences between values.
Robustness: Because it uses ranks, it is less sensitive to extreme outliers than Pearson's correlation.
When to Use: Use Spearman for ordinal data, skewed data, or non-linear monotonic data. 
Handbook of Biological Statistics
Handbook of Biological Statistics
 +5
While Spearman can detect linear relationships (if the data is linear, it is also monotonic), its primary advantage is assessing non-linear, monotonic associations

*** for quadratric no** but for this case** yes
**give example of exponetntial**
**less robust to outliers**

```{r}
x <- c(1, 2, 3, 4, 5, 6, 7, 8, 9, 100)   # outlier at end
y <- c(2, 4, 5, 4, 5, 7, 8, 9, 10, 11)

cor(x, y, method = "pearson")   # pulls low due to outlier
cor(x, y, method = "spearman")  # robust, stays high
```

## Reading

-   P&G Chapter 18

-   OI: Chapter 8

## Back to Lab 2...

```{r}
library(tidyverse)
cdc <- read.csv("https://aggreenbean.github.io/Bios600/labs/data/cdc_cleaned.csv")
```

```{r}
cdc %>%
  ggplot(aes(x = Obesity, y = Exercise)) +
  geom_point() +
  theme_bw() +
  labs(x = "Adequate aerobic activity (%)", 
       y = "Obesity (%)")
```

## Correlation and regression {.smaller}

-   Correlation measures the direction and strength of linear relationships

-   Regression can be used to further quantify these relationships

-   A **regression** line summarizes the relationship between explanatory or predictor variables (e.g., Exercise % or $X$) and response or outcome variables (e.g., Obesity % or $Y$)

-   The fitted regression line can be used to **predict the outcome** for a given set of predictor values. Our predictions do have error, called **residuals** (vertical distance between observation and the regression line).

-   The **least-squares regression line** minimizes the sum of the squared errors.

## Regression

-   In regression, we use one variable (or more) to try to predict values of another.

-   In simple linear regression, one variable $Y$ is the **response** or **outcome** or **dependent variable** and the other $X$ is the **predictor** or **explanatory** variable or **independent** variable.

-   This distinction is critical. The regression of $Y$ on $X$ is **not equal** to the regression of $X$ on $Y$

-   The regression of $Y$ on $X$ can be used to predict $Y$ based on fixed values of $X$.

## Back to Lab 2... {.smaller}

```{r}
cdc %>%
  ggplot(aes(x = Obesity, y = Exercise)) +
  geom_point() +
  theme_bw() +
  labs(x = "Adequate aerobic activity (%)", 
       y = "Obesity (%)") +
  geom_smooth(method = "lm", se = F) 
```

\footnotesize

-   We see the regression line relating exercise and obesity %

-   Points represent actual data values; note that the line is not a perfect fit

## Code for previous slide

```{r}
#| echo: true
#| eval: false
cdc |>
  ggplot(aes(x = Obesity, y = Exercise)) +
  geom_point() +
  theme_bw() +
  labs(x = "Adequate aerobic activity (%)", 
       y = "Obesity (%)") +
  geom_smooth(method = "lm", se = F) 
```


## The linear model

The model is given by $y_i=β_0+β_1x_i+ϵ_i$, where

-   $y_i$ is the outcome (dependent variable) of interest

-   $\beta_0$ is the intercept parameter

-   $\beta_1$ is the slope parameter

-   $x_i$ is the predictor variable

-   $\epsilon_i$ is the error (like $\beta_0$ and $\beta_1$, it is not observed)

The $i$ is an index of observations. So, each observation of the outcome is related to an observation of that individual's value of the predictor.

## The least squares model {.smaller}

Fitted line equation:

$$\hat y_i = \hat \beta_0 +\hat \beta_1 x_i$$

Residual:

$$\hat \epsilon_i = y_i - (\hat \beta_0 +\hat \beta_1 x_i) = y_i - \hat y_i$$

Least squares selects $\hat β_0$ and $\hat β_1$ such that the **error sum of squares**

$$\sum_{i=1}^n \hat \epsilon_i^2 = \sum_{i=1}^n (y_i - \hat y_i)^2$$

is minimized.

## Review slope-intercept form of a line {.smaller}

Recall the slope-intercept form of a line, $y=mx+b$.

-   $b$ is the y-intercept, or where the line crosses the y-axis. It is the predicted value of $y$ when $x = 0$

-   $m$ is the slope, which tells us the **predicted increase** in $y$ when $x$ **changes by 1 unit**

## Review: formula of a line

::: callout-tip

## Review

Consider the line $y=3x+1$. 

- What value does $y$ take when $x=0$?

- As $x$ increases by 1 unit, how much do we expect $y$ to increase by?
:::

-   In statistics, we often represent the **slope** with $β_1$ and the **intercept** with $β_0$, and these values are usually not known but must be estimated.

## Model output {.smaller}

```{r echo = T}
fit <- lm(Obesity ~ Exercise, data = cdc)
summary(fit)
```

## A neater way

```{r}
#| echo: true
library(tidymodels)
library(kableExtra)

tidy(fit) |>
  kable()
```

- Can we do better?


## Even better

```{r}
#| echo: true
tidy(fit) |>
  kable(digits = 3) # round to 3 digits
```

- Note: p-values are not actually 0. They are just quite small so they get rounded to zero in the table. 

::: callout-tip
- If you include a regression model in your project, display the output with `tidy()` and `kable()` like this!
:::

## Intercept and slope

-   The **intercept** is the estimated mean outcome when the predictor is zero (does this make sense in context?)

-   The **slope** is the expected change in the outcome corresponding to a one-unit change in the predictor


**Slope Interpretation**: For each percentage increase in `Exercise`, we expect the `Obesity` percentage to decrease by .48.

::: callout-tip
For all interpretations, remember to **include units**!
:::

## Intercept interpretation

- **The technical intercept interpretation**: When zero percent of the region gets adequate `Exercise`, we would expect the percent of the region that is obese to be 54.79%. 

- Does it make sense to interpret the intercept here?

- You should only interpret the intercept in a regression model if:

  - the predictor can **feasibly** take values equal to or near zero, or
  - there are values near zero in the observed data
  
## Let's check

```{r}
#| echo: true
hist(cdc$Exercise)
```

## Let's check

```{r}
#| echo: true
min(cdc$Exercise)
```

- Since the predictor (`Exercise`) does not feasibly take on values equal to zero (and we also checked that the lowest value in the dataset is 37.4%), we should not interpret the intercept. 

## Model assumptions {.smaller}

Assumptions of the model $y_i = \beta_0 + \beta_1 x_1 + \epsilon_i$ include:

-   The outcomes $y_i$ are independent. This is violated if our study contains repeated outcome measures on an individual, if siblings are enrolled, etc.

-   A linear relationship holds (though we can relax this to some extent)

-   For a specified value of $x$, which is measured without error, $y_i∼N(β_0+β_1x_i,σ^2)$. Note that this implies $ϵ_i∼N(0,σ^2)$.

-   The variance $σ^2$ is constant across all values of $x$; this is called **homogeneity of variance**

## Hypothesis testing {.smaller}

-   The primary hypothesis test of interest is usually $H_0:β_1=0$ (There is no association between exercise and obesity percentage) versus $H_1:β_1≠0$ (there is an association between exercise and obesity percentage.)

-   Under the null hypothesis, we can use a t-test to test this.

- R automatically tests this with the `lm()` function, and provides the t-statistic and p-value in the output.

```{r}
tidy(fit) |>
  kable(digits = 3) # round to 3 digits
```

::: {.callout-tip appearance="simple"}
Given our t-statistic of -7.7 and p-value of \<0.001, what might we conclude in our data?
:::

## Some additional output {.smaller}

```{r}
#| echo: true
summary(fit)$r.squared
```

- The $R-squared$ value indicates that exercise explains 55% of the variance of obesity. 

- $R^2$ tells us about how much of the variability in obesity is explained by exercise; as it gets close to 1, the points get tighter around the regression line.

## Predicted values 

- The regression equation can be used to get predicted means at any value of $x$. 

- $\hat y_i=\hat β_0+ \hat β_1x_i$, where we call $\hat y$ the predicted value of $y$ at a given value of $x$ (what happened to the error term?). 

## Predicted values

- For the CDC data that predicts a state's obesity percentage by its adequate exercise percentage, we have

$$\hat y_i = 54.79 - 0.49 x_i$$

So, the predicted mean obesity percentage for a state where 50% of residents get adequate exercise is

$$\hat y_i = 54.79 - 0.49 (50) $$

We would predict an obesity percentage of approximately 30%. Note the importance of getting the units correct.


## Recap

- What is a linear regression model? How do we fit & visualize the model in R?

- Slope, intercept and interpretations

- Understand when it's appropriate to interpret the intercept

- Interpret $R^2$ from regression output

## Next class

- We will learn how to check the regression model assumptions in R

- Multiple regression
