---
title: "Lecture 20: Logistic regression"
subtitle: "BIOS 600 - Spring 2026"
institute: "UNC - Chapel Hill"
logo: "images/logo_600.png"
format:
  revealjs:
    css: reveal-custom.css
execute:
  freeze: auto
---

## Announcements

- Test corrections due **Thursday 11/20 at 11:59pm** on Canvas. 

- Lab 08 due **Friday 11/21 at 11:59pm**. This is the last lab!

- Lab session today at 3:30pm - swing by to get help from TAs (on project or lab!).

- **Tuesday 11/25**: No class & no lab session - it's a Project Workday. Post in Ed Discussion if you need assistance!

## Announcements

- Project Proposal feedback will be completed by today or tomorrow. 

- If your data is not already in **tidy format** (observations as rows, variables as columns), it will be a lot easier if it is! 


- Reach out if you need help converting your dataset. 

\vskip 3cm

## Overview

- Assessing multicollinearity with VIF

- Logistic regression

- Interpreting estimated coefficients from a logistic regression model

## Computational setup

```{r}
#| echo: true
#| message: false
#| warning: false
library(tidyverse)
library(tidymodels)
library(kableExtra)
library(car) # for VIF
library(MASS)
data(Pima.tr) # for Pima data
library(randomForestSRC) # for pbc data
data(pbc, package = "randomForestSRC")
```

## Very optional reading

-   P&G Chapter 20

-   OI: Section 9.5

## Multicollinearity

```{r}
#| echo: false
#| warning: false
#| message: false
cdc <- read.csv("https://karamccor.github.io/b6002/labs/data/cdc_cleaned.csv")
```

-   One common problem in multiple regression is **multicollinearity**, which occurs when multiple highly correlated variables are used as predictors.

- **Collinearity** simply refers to two variables behind highly correlated.

-   In this case, the model can **become unstable** (often seen as standard errors that get huge and lead to huge confidence interval estimates), and it can be difficult to assess the relationships of the predictors.


## Example: Pima Dataset

Below is an example using the `Pima.tr` dataset from the **MASS** package.  This dataset includes physiological measurements from female Pima Indian patients and is commonly used to study diabetes risk.

- `glu`: plasma glucose concentration
- `skin`: triceps skinfold thickness (mm)
- `bmi`: body mass index (kg/$m^2$)
- `age`: age in years

\footnotesize

```{r}
#| echo: true
mod_pima <- lm(glu ~ bmi + skin + age, data = Pima.tr)
tidy(mod_pima) |>
  kable(digits = 3)
```
## Checking correlation

- Hmm, let's check the correlation matrix:

```{r}
#| echo: true
Pima.tr |>
  dplyr::select(bmi, skin, age) |>
  cor()
```
::: callout-tip

What do you notice?

:::

## Checking VIF

- Another way to check for multicollinearity is through VIF values. VIF values indicate the **extent of multicollinearity**.

- VIF values could range from 1 to infinity. 

  - A VIF value of 1 suggests no correlation with other predictors.

  - Values between 1 and 5 generally indicate **moderate correlation**.

  - Values above 5 (or sometimes 10, depending on the context) are considered problematic, indicating **severe multicollinearity** and potential issues with coefficient stability and interpretation. 


## VIF

- Let's say we had the following model:

```{r}
#| echo: true
fit4 <- lm(Obesity ~ Exercise + Smoking + Over65, 
           data = cdc)
vif_values <- vif(fit4) # car package
vif_values
```
- In this model, none of the VIFs are above 5 (in fact, they're much closer to 1), so multicollinearity is not an issue here. 


# Logistic regression

## Models for binary outcomes

```{r message = F, warning = F}
library(tidyverse)
```

-   Suppose we have a binary outcome variable (e.g., $Y = 1$ if a condition is satisfied and 0 otherwise), and potentially multiple predictors on a variety of scales.

::: {.callout-tip appearance="simple"}
What distribution does our outcome have?
:::

## Models for binary outcomes

-   Suppose we want to model $P(Y=1)$, the success probability. One strategy might be to simply fit a linear regression model to the probabilities:

$$P(Y_i = 1) = \beta_0 + \beta_1 x_{i1} + \cdots + \beta_p x_{ip} + \epsilon_i$$

::: {.callout-tip appearance="simple"}
Is this a reasonable approach? Why or why not?
:::

## Primary biliary cirrhosis

The Mayo Clinic conducted a trial for primary biliary cirrhosis, comparing the drug D-penicillamine vs. placebo. Patients were followed for a specified duration, and their status at the end of follow-up (whether they died) was recorded. Researchers are interested in predicting whether a patient died based on the following variables:

-   `ascites`: whether the patient had ascites (abnormal fluid buildup in the abdomen) (1 = yes, 0 = no)

-   `bili`: serum bilirubin in mg/dL (higher values may indicate liver damage/disease)

-   `stage`: histologic stage of disease (ordinal categorical variable with stages 1, 2, 3, and 4)

## What can go wrong?

Suppose we fit the following model:

$$P(Y_i = 1) = \beta_0 + \beta_1 asc_i + \beta_2 bili_i + \beta_3 s2_i + \beta_4 s3_i + \beta_5 s4_i + \epsilon_i$$

where $Y=1$ if a patient died during the study follow-up.

::: {.callout-tip appearance="simple"}
What can go wrong?
:::

## What could go wrong?

- We *can* technically fit the model:
```{r}
#| echo: true
model <- lm(status ~ ascites + bili + as.factor(stage), 
            data = pbc)
tidy(model) |>
  kable(digits = 3)
```

- However, let's check the fitted vs. residuals plot.

## What could go wrong?

```{r}
ggplot(model, aes(x = .fitted, y = .resid)) +
  geom_point() +
  geom_hline(yintercept = 0, col = "red") +
  labs(xlab = "Fitted", 
       ylab = "Residuals")
```

## From probabilities to log odds

-   Suppose the probability of an event is $p$.

-   Then the **odds** that the event occurs is $\frac{p}{1-p}$

-   Taking the natural log of the odds, we have the **logit** of $p$ (also known as the **log-odds**):

::: poll
$$logit(p) = \text{log} \left(\frac{p}{1-p}\right)$$
:::

-   Although $p$ is constrained to lie between 0 and 1, the logit of $p$ can be anything from $-\infty$ to $\infty$.

## Logistic regression model

Let's create a linear model for the log-odds:

::: poll
$$logit(p_i) = \text{log} \left(\frac{p_i}{1-p_i}\right) = \beta_0 + 
\beta_1 x_{i1} + \cdots + \beta_p x_{ip}$$
:::

- This is the **logistic regression** model. 

- In this model, it's ok that we could get any potential value for the logit of $p$  (can range from $-\infty$ to $+ \infty$). 

- Once we get our $\hat\beta$ estimates, we can always back-transform to get a valid probability between 0 and 1.

## Interpreting parameters

-   Suppose that X is a binary (0/1) variable (e.g., X = 1 for smoker and 0 for non-smoker). In this case, we interpret $\text{exp}(\beta_1)$ as the **odds ratio** (OR) of the response for the two possible levels of X.

-   For X on a continuous scale, $\text{exp}(\beta_1)$ is interpreted as the odds ratio of the response comparing two values of X one unit apart.

## Back to PBC

```{r}
#| echo: true
fit2 <- glm(status ~ ascites + bili + as.factor(stage), 
            data = pbc, 
            family = "binomial")
tidy(fit2) |>
  kable(digits = 3)
```

## Interpretations

-   $\hat\beta_{\text{ascites}}$: The estimate for ascites was 2.87. Thus, the odds ratio for dying is exp(2.87), or approx. 17.6. Patients with ascites have 17.6 times the odds of dying compared to patients that do not, holding all other variables constant.

- $\hat\beta_{\text{bili}}$: The estimate for bilirubin was 0.31. Thus, the odds ratio for dying for a patient with 1 additional mg/dL serum bilirubin compared to another is exp(0.31) ≈ 1.36, holding all other variables constant.

-   $\hat\beta_{\text{stage 3}}$: The baseline stage was 1. The estimate corresponding to stage 3 was 1.72. Thus, patients in stage 3 have approximately 5.58 times the odds of dying compared to patients with the baseline stage, holding all other variables constant.

::: callout-tip
## Your turn

What is the interpretation for $\hat\beta_{\text{stage 4}}$

:::



## Logistic regression model

::: poll
$$logit(p_i) = \text{log} \left(\frac{p_i}{1-p_i}\right) = \beta_0 + 
\beta_1 x_{i1} + \cdots + \beta_p x_{ip}$$
:::

- Negative logits represent probabilities less than one-half. 

- Positive logits represent probabilities above one-half.

## Interpreting parameters

Typically we interpret *functions* of parameters in logistic regression rather than the parameters themselves. For the simple model

::: poll
$$\text{log} \left(\frac{p_i}{1-p_i}\right) = \beta_0 + 
\beta_1 x_{i}$$
:::

then the probability that $Y = 1$ when $X = 0$ is

$\frac{\text{exp}(\beta_0)}{1 + \text{exp}(\beta_0)}$

- How did we get this?

::: callout-warning


If you're curious how to convert logit to probability, see next slide. Prepare for math ahead! (This won't be on the exam!)

:::

## Solving logistic model for $p_i$

\footnotesize

$$\text{log} \left(\frac{p_i}{1-p_i}\right) = \beta_0 + 
\beta_1 x_{i}$$

$$\left(\frac{p_i}{1-p_i}\right) = \text{exp}(\beta_0 + 
\beta_1 x_{i})$$

$$p_i = (1-p_i)\text{exp}(\beta_0 + 
\beta_1 x_{i})$$

$$p_i = \text{exp}(\beta_0 + 
\beta_1 x_{i}) - p_i \text{exp}(\beta_0 + 
\beta_1 x_{i}) $$

$$p_i + p_i \text{exp}(\beta_0 + 
\beta_1 x_{i})  = \text{exp}(\beta_0 + 
\beta_1 x_{i}) $$

$$p_i(1 + \text{exp}(\beta_0 + 
\beta_1 x_{i}))  = \text{exp}(\beta_0 + 
\beta_1 x_{i}) $$

$$p_i = \frac{\text{exp}(\beta_0 + 
\beta_1 x_{i})}{1 + \text{exp}(\beta_0 + 
\beta_1 x_{i}) } $$





## Interpreting parameters {.smaller}

-   Why is this the interpretation for continuous variables?

-   The log odds of response for X = 1 is given by $β_0+β_1$, and the log odds of response for X = 0 is $β_0$.

-   So the odds ratio of response comparing X = 1 to X = 0 is given by:

$$\frac{\text{exp}(\beta_0 + \beta_1)}{\text{exp}(\beta_0)} = \text{exp}(\beta_1)$$

-   In a multiple logistic regression model, this has to be interpreted conditionally on values of other variables in the model (i.e., controlling for them).



## Back to PBC {.smaller}

Remember, we are interested in the probability that a patient died during follow-up (a "success”). We are predicting the log-odds of this event happening.



## Predicted probabilities {.smaller}

-   There is a one-to-one-relationship between $p$ and $logit(p)$. So, if we predict the log-odds, we can back-transform to get back to a predicted probability.

-   For instance, suppose a patient does not have ascites, has a bilirubin level of 5 mg/dL, and is a stage 2 patient. Their predicted log-odds are:

$$−3.14+0.31×5+1.25=−0.34$$

Thus, their predicted probability of dying is

$$\frac{\text{exp}(-0.34)}{1+\text{exp}(-0.34)} = 0.42$$


## On your own

::: {.callout-tip appearance="simple"}
What is the predicted probability of dying for a patient with ascites, a bilirubin level of 3 mg/dL, and stage 3?
:::

## Hypothesis tests in logistic regression

Generally, we wish to know whether the OR = 1 or equivalently, whether the logit of $p$ (a $\beta$ coefficient) = 0. To test

$$H_0: \beta_k = 0$$

$$H_1: \beta_k \neq 0$$

we can compare the ratio of a parameter estimate to its standard error using the standard normal distribution (reason we use Z instead of t is a bit technical).

## Confidence intervals

CIs for associations on the **logit scale** are given by

::: poll
$$\hat \beta_j \pm z^*_{1-\alpha/2} \times \widehat {SE} (\hat\beta_j)$$
:::

These are typically translated into confidence intervals in **odds scale** by exponentiating the lower and upper limits.

## Next class

- Application Exercise, bring your computer!
