---
title: "Lecture 21: Logistic Regression Continued"
subtitle: "BIOS 600 - Spring 2026"
institute: "UNC - Chapel Hill"
logo: "images/logo_600.png"
format:
  revealjs:
    css: reveal-custom.css
execute:
  freeze: auto
---

## Announcements

- Test corrections due **Friday** at 11:59 pm


## Announcements

- Final Exam is **Tuesday, May 5, 8am-11am**
  - Will be in this room **and** McGavern 2301. 

- Final Exam will be similar length to previous exams

- Focus on material since Exam 02

- Please reach out **ASAP** if you have any scheduling issues

## Course evaluation

- Course evaluations are due **April 28 at 11:59pm**.

- Please fill these out

- If at least 85% of students fill out the course evaluations, everyone in the class will receive an extra 1/2 point on the final exam score. 


## Computational setup

```{r}
#| echo: true
#| message: false
#| warning: false
library(tidyverse)
library(tidymodels)
library(kableExtra)
library(randomForestSRC) # for pbc data
data(pbc, package = "randomForestSRC")
```

## When to choose what method? {.smaller}

| Predictor Type      | Outcome Type           | Common Tests / Topics                             | 
|----------------------|------------------------|---------------------------------------------------|
| Categorical          | Categorical            | Fisher’s exact test, $\chi^2$ test                      |
| Categorical          | Continuous             | t-tests, ANOVA, nonparametric alternatives        |
| Continuous           | Continuous             | Correlation                   |
| Continuous **or** Categorical           | Continuous             | Regression                          |
| Continuous **or** Categorical           | Categorical            | Logistic regression              | 
| Other / Complex      | Various (e.g. survival, counts) | Advanced or “exotic” methods               |

## Quick review

::: {.callout-tip}

## Review

1. Which R function do we use to run a linear regression? How about for a logistic regression?

2. What additional piece of information do you need to specify when running a logistic regression, compared to a linear regression?

3. Suppose you have an outcome variable $Y$ that takes on one of two values, 0 or 1. What distribution does $Y_i$ have (one observation)? How about $Y_1, \ldots, Y_n$ ($n$ observations)?

4. What is the logit function? (describe in words & symbols)

5. (Fill in the blank:) Negative logits correspond to probabilities less than ______.

:::

## Logistic regression from last time

Recall the cirrhosis dataset from last class. 

Researchers are interested in predicting whether a patient died (`status`) based on the following variables:

-   `ascites`: whether the patient had ascites (abnormal fluid buildup in the abdomen) (1 = yes, 0 = no)

-   `bili`: serum bilirubin in mg/dL (higher values may indicate liver damage/disease)

-   `stage`: histologic stage of disease (ordinal categorical variable with stages 1, 2, 3, and 4)

## Logistic regression model {.smaller}

```{r}
#| echo: true
fit2 <- glm(status ~ ascites + bili + as.factor(stage), 
            data = pbc, 
            family = "binomial")
tidy(fit2) |>
  kable(digits = 3)
```

## Hypothesis tests in logistic regression

Generally, we wish to know whether the OR = 1 or equivalently, whether the logit of $p$ (a $\beta$ coefficient) = 0. To test

$$H_0: \beta_k = 0$$

$$H_1: \beta_k \neq 0$$

we can compare the ratio of a parameter estimate to its standard error using the standard normal distribution (reason we use Z instead of t is a bit technical).

## Logistic regression model {.smaller}

```{r}
#| echo: true
fit2 <- glm(status ~ ascites + bili + as.factor(stage), 
            data = pbc, 
            family = "binomial")
# Gets estimates for log odds scale
tidy(fit2) |>
  kable(digits = 3)
```

## Interpretation

$\hat\beta_{\text{bili}}$: Each one mg/dL increase in bilirubin is associated with the odds of death being multiplied by exp(.315) = 1.37, holding ascites presence and ascites presence constant.

**Another Equivalent Interpretation**: A one mg/dL increase in bilirubin is associated with 37% higher odds of death, holding ascites presence and status constant.

Interpretation includes **estimates of magnitude, the units and since we are using multivariable model specifies we are holding all other varibles constant**

## In R {.smaller}

\footnotesize

```{r}
#| echo: true
fit2 <- glm(status ~ ascites + bili + as.factor(stage), 
            data = pbc, 
            family = "binomial")
# Get odds ratios 
# Note std.error values are **not** not exponentiated
tidy(fit2, exponentiate = TRUE) |>
  kable(digits = 3)
```

## Confidence intervals

CIs for associations on the **logit scale** are given by

::: poll
$$\hat \beta_j \pm z^*_{1-\alpha/2} \times \widehat {SE} (\hat\beta_j)$$
:::

These are typically translated into confidence intervals in **odds scale** by exponentiating the lower and upper limits.

$$[\text{exp}(\hat \beta_j - z^*_{1-\alpha/2} \times \widehat {SE} (\hat\beta_j)), \text{exp}(\hat \beta_j + z^*_{1-\alpha/2} \times \widehat {SE} (\hat\beta_j))]$$

## Confidence intervals

- For the variable ascites [patient had ascites (1 = yes, 0 = no)]
  - $\hat{\beta}_{ascites}$ = 2.868
  - Standard error of $\hat{\beta}_{ascites}$ = 1.067

- We are 95% confident that the true difference in the log‑odds of the outcome comparing patients with ascites to those without ascites lies within the following interval, holding the other variables in the model constant: $$\hat \beta_j \pm z^*_{1-\alpha/2} \times \widehat {SE} (\hat\beta_j)$$

## In R {.smaller}

\footnotesize

```{r}
#| echo: true
# Get odds ratios and 95% CIs
tidy(fit2, conf.int = TRUE) |>
  kable(digits = 3)
```

## Confidence intervals

- For the variable ascites [patient had ascites (1 = yes, 0 = no)]
  - $\hat{\beta}_{ascites}$ = 2.868
  - Standard error of $\hat{\beta}_{ascites}$ = 1.067
  
- We are 95% confident that the true odds ratio comparing patients with ascites to those without ascites lies within the following interval, holding the other variables in the model constant: $$[\text{exp}(\hat \beta_j - z^*_{1-\alpha/2} \times \widehat {SE} (\hat\beta_j)), \text{exp}(\hat \beta_j + z^*_{1-\alpha/2} \times \widehat {SE} (\hat\beta_j))]$$

## In R {.smaller}

\footnotesize

```{r}
#| echo: true
# Get odds ratios and 95% CIs
tidy(fit2, exponentiate = TRUE, conf.int = TRUE) |>
  kable(digits = 3)
```

## Predicted probabilities {.smaller}

-   There is a one-to-one-relationship between $p$ and $logit(p)$. So, if we predict the log-odds, we can back-transform to get back to a predicted probability.

-   For instance, suppose a patient does not have ascites, has a bilirubin level of 5 mg/dL, and is a stage 2 patient. Their predicted **log-odds** are:

$$−3.14+0.31×5+1.25=−0.34$$

Thus, their predicted **probability** of dying is

$$\frac{\text{exp}(-0.34)}{1+\text{exp}(-0.34)} = 0.42$$


## In R

\footnotesize

```{r}
#| echo: true
# Example patient:
new_patient <- data.frame(
  ascites = 0,
  bili = 5,
  stage = 2
)

# Get predicted log-odds (type = "link")
pred_logodds <- predict(fit2, 
                        newdata = new_patient, 
                        type = "link")

# Convert log-odds to odds
pred_odds <- exp(pred_logodds)

# Convert odds to probability
pred_prob <- pred_odds / (1 + pred_odds)
```

## In R

- Type = link tells R to estimate the outcome variable in terms of a function of the linear predictor (in the case of logistic regression this returns the log-odds (logit) of the outcome)

- Since generalized linear models do not model the outcome directly in the **glm** function we need to specify the distribution via the family argument 

- Since generalized linear models do not model the outcome directly in the **predict** function we need to specify the scale on which predictions are returned (e.g. log‑odds vs. probabilities)

## Output

```{r}
#| echo: true
pred_logodds
pred_prob
```

## Alternatively...

- We can also calculate the predicted probability directly using `predict` by setting the `type` to be `"response"`. 

```{r}
#| echo: true
# Get predicted probability (type = "response")
pred_prob2 <- predict(fit2, 
                        newdata = new_patient, 
                        type = "response")
pred_prob2
```

## Application Exercise

::: callout-tip

Head to Canvas and download the template for Application Exercise 05. AE 05 will be due **this Sunday at 11:59pm**.

Selected groups should put their answers (screenshots of code, interpretations) on [these google slides](https://docs.google.com/presentation/d/1Jm7LUn5_xbzZDmswAJQmoH2M9rsnYx68-ui_Vqlx7Fw/edit?usp=sharing). Pick a representative to present your slide.

:::

## More practice

::: {.callout-tip appearance="simple"}

```{r}
tidy(fit2) |>
  kable(digits = 3)
```

By hand, write out the following. Leave in terms of values given:

1. What is the **predicted log odds** of dying for a patient with ascites, a bilirubin level of 3 mg/dL, and stage 3?

2. Given this value, what is the **predicted probability**?

:::

## Hypothesis tests in logistic regression

Generally, we wish to know whether the OR = 1 or equivalently, whether the logit of $p$ (a $\beta$ coefficient) = 0. To test

$$H_0: \beta_k = 0$$

$$H_1: \beta_k \neq 0$$

we can compare the ratio of a parameter estimate to its standard error using the standard normal distribution 

- (The reason we use Z instead of t is a bit technical, so we won't go into it in class).

## Confidence intervals

CIs for associations on the **logit scale** are given by

::: poll
$$\hat \beta_j \pm z^*_{1-\alpha/2} \times \widehat {SE} (\hat\beta_j)$$
:::

These are typically translated into confidence intervals in **odds scale** by exponentiating the lower and upper limits. That is,:

$$[\text{exp}(\hat \beta_j - z^*_{1-\alpha/2} \times \widehat {SE} (\hat\beta_j)), \text{exp}(\hat \beta_j + z^*_{1-\alpha/2} \times \widehat {SE} (\hat\beta_j))]$$

## Participation

- Think about your knowledge of statistics and confidence with R coding at the beginning of the semester vs. now. 

- What (if anything) about the structure of our class has been particularly helpful for your learning? (labs, Application Exercises, HW, lecture examples, TA office hours?)

## Recap

- Logistic regression review

- Interpreting continuous and categorical coefficients in a logistic regression model

- Predicting probabilities using a logistic regression model object

## Next class

- No class on Tuesday 11/25. 

- Have a restful week next week! 

- Last day of class is Dec. 2. (Class wrap-up, Exam info)
