---
title: "Lecture 21: Logistic Regression Continued"
subtitle: "BIOS 600 - Spring 2026"
institute: "UNC - Chapel Hill"
logo: "images/logo_600.png"
format:
  revealjs:
    css: reveal-custom.css
execute:
  freeze: auto
---

## Announcements

- Test corrections due **tonight at 11:59pm**

- Lab 08 due Friday at 11:59pm

- No class/lab on **Tuesday, November 25**. It's a project workday.

- Projects due **Monday, December 1 at 11:59pm**
  - PDF report and Quarto file (when rendered, quarto file should produce PDF report exactly), slides to be turned in on Canvas 
  - Video to be posted on Canvas > Panopto

- Do not wait until the last minute to complete your projects! 

## Announcements

- Final Exam is **Thursday, Dec 11, 8am-11am** in this room (MHRC 0001). 

- Final Exam will be similar length to previous exams, and will focus on material since Exam 02. 

## Course evaluation

- Please fill out the course evaluations, due **December 04 at 11:59pm**.

- If at least 85% of students fill out the course evaluations, everyone in the class will receive an extra 1/2 point on the final exam score. 


## Computational setup

```{r}
#| echo: true
#| message: false
#| warning: false
library(tidyverse)
library(tidymodels)
library(kableExtra)
library(randomForestSRC) # for pbc data
data(pbc, package = "randomForestSRC")
```



## Quick review

::: {.callout-tip}

## Review

1. Which R function do we use to run a linear regression? How about for a logistic regression?

2. What additional piece of information do you need to specify when running a logistic regression, compared to a linear regression?

3. Suppose you have an outcome variable $Y$ that takes on one of two values, 0 or 1. What distribution does $Y_i$ have (one observation)? How about $Y_1, \ldots, Y_n$ ($n$ observations)?

4. What is the logit function? (describe in words & symbols)

5. (Fill in the blank:) Negative logits correspond to probabilities less than ______.

:::

## Logistic regression from last time

Recall the cirrhosis dataset from last class. 

Researchers are interested in predicting whether a patient died (`status`) based on the following variables:

-   `ascites`: whether the patient had ascites (abnormal fluid buildup in the abdomen) (1 = yes, 0 = no)

-   `bili`: serum bilirubin in mg/dL (higher values may indicate liver damage/disease)

-   `stage`: histologic stage of disease (ordinal categorical variable with stages 1, 2, 3, and 4)

## Logistic regression model

```{r}
#| echo: true
fit2 <- glm(status ~ ascites + bili + as.factor(stage), 
            data = pbc, 
            family = "binomial")
tidy(fit2) |>
  kable(digits = 3)
```

## Interpretation

$\hat\beta_{\text{bili}}$: Each one mg/dL increase in bilirubin is associated with the odds of death being multiplied by exp(.315) = 1.37, holding ascites presence and ascites presence constant.

**Another Equivalent Interpretation**: A one mg/dL increase in bilirubin is associated with 37% higher odds of death, holding ascites presence and status constant.

## Predicted probabilities {.smaller}

-   There is a one-to-one-relationship between $p$ and $logit(p)$. So, if we predict the log-odds, we can back-transform to get back to a predicted probability.

-   For instance, suppose a patient does not have ascites, has a bilirubin level of 5 mg/dL, and is a stage 2 patient. Their predicted **log-odds** are:

$$−3.14+0.31×5+1.25=−0.34$$

Thus, their predicted **probability** of dying is

$$\frac{\text{exp}(-0.34)}{1+\text{exp}(-0.34)} = 0.42$$

## In R

\footnotesize

```{r}
#| echo: true
# Example patient:
new_patient <- data.frame(
  ascites = 0,
  bili = 5,
  stage = 2
)

# Get predicted log-odds (type = "link")
pred_logodds <- predict(fit2, 
                        newdata = new_patient, 
                        type = "link")

# Convert log-odds to odds
pred_odds <- exp(pred_logodds)

# Convert odds to probability
pred_prob <- pred_odds / (1 + pred_odds)
```

## Output

```{r}
#| echo: true
pred_logodds
pred_prob
```

## Alternatively...

- We can also calculate the predicted probability directly using `predict` by setting the `type` to be `"response"`. 

```{r}
# Get predicted probability (type = "response")
pred_prob2 <- predict(fit2, 
                        newdata = new_patient, 
                        type = "response")
pred_prob2
```

## Application Exercise

::: callout-tip

Head to Canvas and download the template for Application Exercise 05. AE 05 will be due **this Sunday at 11:59pm**.

Selected groups should put their answers (screenshots of code, interpretations) on [these google slides](https://docs.google.com/presentation/d/1Jm7LUn5_xbzZDmswAJQmoH2M9rsnYx68-ui_Vqlx7Fw/edit?usp=sharing). Pick a representative to present your slide.

:::

## More practice

::: {.callout-tip appearance="simple"}

```{r}
tidy(fit2) |>
  kable(digits = 3)
```

By hand, write out the following. Leave in terms of values given:

1. What is the **predicted log odds** of dying for a patient with ascites, a bilirubin level of 3 mg/dL, and stage 3?

2. Given this value, what is the **predicted probability**?

:::

## Hypothesis tests in logistic regression

Generally, we wish to know whether the OR = 1 or equivalently, whether the logit of $p$ (a $\beta$ coefficient) = 0. To test

$$H_0: \beta_k = 0$$

$$H_1: \beta_k \neq 0$$

we can compare the ratio of a parameter estimate to its standard error using the standard normal distribution 

- (The reason we use Z instead of t is a bit technical, so we won't go into it in class).

## Confidence intervals

CIs for associations on the **logit scale** are given by

::: poll
$$\hat \beta_j \pm z^*_{1-\alpha/2} \times \widehat {SE} (\hat\beta_j)$$
:::

These are typically translated into confidence intervals in **odds scale** by exponentiating the lower and upper limits. That is,:

$$[\text{exp}(\hat \beta_j - z^*_{1-\alpha/2} \times \widehat {SE} (\hat\beta_j)), \text{exp}(\hat \beta_j + z^*_{1-\alpha/2} \times \widehat {SE} (\hat\beta_j))]$$

## Participation

- Think about your knowledge of statistics and confidence with R coding at the beginning of the semester vs. now. 

- What (if anything) about the structure of our class has been particularly helpful for your learning? (labs, Application Exercises, HW, lecture examples, TA office hours?)

## Recap

- Logistic regression review

- Interpreting continuous and categorical coefficients in a logistic regression model

- Predicting probabilities using a logistic regression model object

## Next class

- No class on Tuesday 11/25. 

- Have a restful week next week! 

- Last day of class is Dec. 2. (Class wrap-up, Exam info)
